Synthesising Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-10-17 papers/termination/termination-esop/paper.tex
added a section on solving the second order SAT formula and cited kalashnikov

\section{Satisfiability of a Synthesis Fragment Formula} \label{sec:synthesis}

In this section we present a solver for the synthesis fragment:
%
\[
  \exists P,~ x . ~\forall~ y . \sigma(P, x, y)
\]
%
where $P$ ranges over functions, and $x$ and $y$ range over ground terms
(as mentioned, we drop vector notation and write $x$ for $\vec{x}$).

By viewing $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times Y^m  \to
\mathbb{B}$ as a specification function that returns true iff $P$ computes
appropriate outputs for the inputs $x$ and $y$, we can regard the
satisfaction problem for the synthesis fragment as \emph{program synthesis},
i.e.~the mechanised construction of software that provably satisfies a
given specification.  A witness $P$ to the satisfiability of some
formula is then just a program that meets the specification $\sigma$.

Research in the area of program synthesis has been
particularly fruitful beginning with Alonzo Church's work on the
\emph{Circuit Synthesis Problem} in the sixties~\cite{church-synth}, and
continuing with works such as {\sc Brahma}~\cite{brahma} and program
sketching~\cite{lezama-thesis,sketch,modular-sketch}.

In the context of termination analysis, a witness to the
satisfiability of a generalised termination specification is a
proof of \mbox{(non-)termination}.  As mentioned earlier in the paper, our
analysis must ensure precise treatment of bit-wise operations and arithmetic
modulo fixed widths by modelling bit-accurate semantics.  The synthesiser
must therefore be able to compute solutions over machine
integers and floating-point numbers from the specifications detailed in
Section~\ref{sec:second.order}.  In the next section, we will begin our treatment of
bit-accurate synthesis by illustrating some of the differences between
machine arithmetic and conventional integer arithmetic.

\begin{theorem}
 \label{thm:synth-semicomplete}
 If the existential first-order theory used to express the specification $\sigma$ is decidable
 and the domain of inputs $X$ is finite,
 the second-order SAT decision procedure is semi-complete -- if a solution $(P, x_0) \models \sigma$
 exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
 satisfies the specification, the algorithm may not terminate.
\end{theorem}

%% \subsection{The Synthesis Specification}

%% The logic we use to express the synthesis formula, i.e.
%% the generalised termination specification, is a subset of C that we call
%% \newC.  The characteristic property of a \newC program is that safety can be
%% decided for it using a single query to a Bounded Model Checker.  A~\newC
%% program is just a C program with the following syntactic restrictions:
%%  all loops in the program must have a constant bound;
%%  all recursion in the program must be limited to a constant depth;
%%  all arrays must be statically allocated (i.e.~not using \texttt{malloc}),
%%  and be of constant size.
%% Additionally, \newC programs may use nondeterministic values, assumptions
%% and arbitrary-width types.

\subsection{The Proof Language}

The proofs generated by the synthesiser take the form of imperative programs
written in the language $\mathcal{L}$.  An $\mathcal{L}$-program is a list
of instructions, each of which matches one of the patterns shown in
Figure~\ref{fig:l-language}.  An instruction has an opcode (such as
\verb|add| for addition) and one or more operands.  An operand is either a
constant, one of the program's inputs or the result of a previous
instruction.  The $\mathcal{L}$ language has various arithmetic and logical
operations, as well as basic branching in the form of the \verb|ite|
(if-then-else) instruction.

% The (non-)termination proofs are arbitrary expressions over the grammar given in Figure~\ref{fig:l-language}.  
% %written in a simple RISC-like language $\mathcal{L}$, whose syntax is given in Fig.~\ref{fig:l-language}.  
% \todo{update the figure and make it look more like a grammar.}

%% \begin{figure}
%%  \begin{center}
%%  \resizebox{\linewidth}{!}{%
%%   \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,
%%  semithick, initial text=]

%%   \matrix[nodes={draw, fill=none, shape=rectangle, minimum height=1cm, minimum width=1.5cm},
%%           row sep=2cm, column sep=2.5cm, ampersand replacement=\&] {
%%    \node (synth) {Synthesise};
%%    \&
%%    \node (verif) {Verify}; %\\
%%    %\node[draw=none] {};
%%    \&
%%    \node[ellipse] (done) {Done}; \\
%%   };

%%    \path
%%     (synth) edge [bend left] node {Candidate program} (verif)
%%     (verif) edge [bend left] node {Counterexample input} (synth)
%%     (verif) edge node {Valid} (done);
%%  \end{tikzpicture}
%%  }
%%  \end{center}

%%  \caption{Abstract synthesis refinement loop
%%  \label{fig:abstract-refinement}}
%% \end{figure}

@@

%% \subsection{The Synthesis Algorithm}

%% We use Counterexample Guided Inductive Synthesis
%% (CEGIS)~\cite{lezama-thesis,sketch,DBLP:conf/iclp/BrainCVF06} to find an
%% $\mathcal{L}$-program satisfying our specification.  The core of the CEGIS
%% algorithm is the refinement loop shown in Figure~\ref{fig:abstract-refinement}
%% and detailed in Algorithm~\ref{alg:cegis}.

%% %\begin{figure}
%% \begin{algorithm}
%% {\small
%%  \caption{Abstract synthesis algorithm
%%  \label{alg:cegis}}
%%  %\vspace{-1.7em}
%%  %\begin{multicols}{2}
%%  \begin{algorithmic}[1]
%%  \Statex
%% \Function{synth}{{\sc inputs}, $\sigma$}
%%   \Let{$(i_1, \ldots, i_N)$}{{\sc inputs}}
%%   \Let{query}{$\exists P, x_0 . \forall i \in \text{\sc inputs} . \sigma(P, x_0, i)$}
%%   \Let{result}{decide(query)}
%%   \If{result.satisfiable}
%%     \State \Return{result.model}
%%   \Else
%%     \State \Return{unsatisfiable}
%%   \EndIf
%% \EndFunction
%% \Statex
%% \Function{verif}{$P, x_0, \sigma$}
%%   \Let{query}{$\exists x . \lnot \sigma(P, x_0, x)$}
%%   \Let{result}{decide(query)}
%%   \If{result.satisfiable}
%%     \State \Return{result.model}
%%   \Else
%%     \State \Return{valid}
%%   \EndIf
%% \EndFunction
%% %\columnbreak
%% \Statex
%% \Function{Synthesise}{$\sigma$}
%%   \Let{{\sc inputs}}{$\emptyset$}
%%   \Loop
%%     \Let{candidate}{\Call{synth}{{\sc inputs}}}
%%     \If{candidate = UNSAT}
%%       \State \Return{unsatisfiable}
%%     \EndIf
%%     \Let{res}{\Call{verif}{candidate}}
%%     \If{res = valid}
%%       \State \Return{candidate}
%%     \Else
%%       \Let{{\sc inputs}}{{\sc inputs} $\cup$ res}
%%     \EndIf
%%   \EndLoop
%% \EndFunction
%%  \end{algorithmic}
%%  %\end{multicols}
%%  }
%% \end{algorithm}
%% %\end{figure}

%% The algorithm is divided into two procedures: {\sc synth} (see Figure~\ref{fig:synth-dfd}) and {\sc
%% verif}, which interact via a finite set of test vectors {\sc inputs}.

%% The {\sc synth} procedure tries to find existential witnesses $(P, x_0)$
%% that satisfy the partial specification:
%% %
%% \[
%%  \exists P, x_0 . \forall x \in \text{\sc inputs} . \sigma(P, x_0, x)
%% \]

%% If {\sc synth} succeeds in finding a witness $(P, x_0)$, this witness is a
%% candidate solution to the full synthesis formula.  We pass this candidate
%% solution to {\sc verif} which determines whether it does satisfy
%% the specification on all inputs by checking satisfiability of the
%% verification formula:
%% %
%% \[
%%  \exists x . \lnot \sigma(P, x_0, x)
%% \]

%% If this formula is unsatisfiable, the candidate solution is in fact a
%% solution to the synthesis formula and so the algorithm terminates. 
%% Otherwise, the witness $x$ is an input on which the candidate solution fails
%% to meet the specification.  This witness $x$ is added to the {\sc inputs}
%% set and the loop iterates again.

%% Concretely, {\sc synth} is implemented as shown in
%% Figure~\ref{fig:synth-dfd}.  We start by taking the termination
%% specification and generating a \newC program which takes as input a candidate
%% $(P, x_0)$ and asserts that the candidate fails to meet the specification on
%% at least one of the elements of {\sc inputs}.  So this program is unsafe iff
%% there is some candidate $(P, x_0)$ satisfying the specification for all the
%% elements of {\sc inputs}.  Finding a new candidate solution is therefore
%% reduced to model checking this \newC program, which by construction is loop
%% free.  The model checking is done using a combination of symbolic Bounded Model
%% Checking (BMC), explicit state enumeration and genetic programming (GP) running in
%% parallel.
%% Finally, if any of the model checkers finds a candidate solution, it returns it
%% and the others are killed.  The form of
%% this candidate solution is yet another program, this time written in the
%% proof language $\mathcal{L}$.

%% % As well as efficiency, compiling the original source code of the program
%% % under analysis has the advantage of guaranteeing that we are verifying the
%% % exact semantics of the program as understood by a compiler, rather than some
%% % ad-hoc abstraction of its semantics.

%% Similarly to {\sc synth}, the {\sc verif} procedure is implemented by
%% creating a \newC program that is unsafe iff some input exists on which the
%% candidate solution fails to meet the specification.  Again, symbolic and
%% explicit-state model checking are used to decide the satisfiability of the
%% verification formula\footnote{GP is not used, since it is inherently incomplete.}.

%% \begin{figure}
%% \begin{center}
%% \tikzstyle{file}=[draw, text width=7.0em, text centered,
%%   minimum height=1.5em]
%% \tikzstyle{process} = [draw, minimum height=3em, circle]
%% \tikzstyle{line} = [draw, color=black, -latex']

%% \def\Divide#1#2{%
%%  \coordinate(a) at ($(#1.east) !.5! (#2.west)$);
%%  \coordinate(b) at (a |- 0,-3);
%%  \draw[dotted] (b) -- ++(0, 4.5);
%% }

%% \resizebox{\linewidth}{!}{
%% \begin{tikzpicture}[font=\sffamily]

%% \node [file] (pua) {input.c};

%% \path (pua.east)+(2,0) node [file] (spec) {\sc Termination Specification};
%% \path (spec.south)+(0.0, -1) node [file] (inputs) {\sc Tracked Inputs};
%% %% \path (synth.south)+(0.0, -0.5) node [file] (tests) {\sc tests.c};
%% %% \path (tests.south)+(0.0, -0.5) node [file] (interpreter) {\sc interpreter.c};
%% %% \path (interpreter.south)+(0.0, -0.5) node [file] (spec) {\sc spec.c};

%% %% \path (spec.east)+(2.0, -0.25) node [process] (merged) {merge};
%% \path (spec.east)+(2.0, -0.75) node [file] (file) {synth.c};

%% \path (file.east)+(2.0, 1.5) node [process] (cbmc) {\sc cbmc};
%% \path (file.east)+(2.0, 0.0) node [process] (gcc) {\sc Search};
%% \path (file.east)+(2.0, -1.5) node [process] (gp) {\sc GP};

%% \path (gcc.east)+(2.5, 0.0) node [file] (out) {Candidate Proof};

%% %\path [line] (spec) -- (merged);

%% \path [line] (pua) -- (spec);

%% \path [line] (spec) -- (file);
%% \path [line] (inputs) -- (file);

%% \path [line] (file) -- (cbmc);
%% \path [line] (file) -- (gcc);
%% \path [line] (file) -- (gp);

%% \path [line] (cbmc) -- (out);
%% \path [line] (gcc) -- (out);
%% \path [line] (gp) -- (out);

%% \Divide{pua}{spec}{puaspec}
%% %\Divide{spec}{file}{specfile}
%% \Divide{file}{gcc}{filegcc}
%% \Divide{gcc}{out}{gccout}

%% \draw (pua |- 0,-3) node [align=center] {Input program \\ (written in C)};

%% \coordinate (midspec) at ($(spec) !.5! (file)$);
%% \draw (midspec |- 0,-3) node [align=center] {Automatically generated specification \\ (written in C)};
%% %\draw (file |- 0,-3) node {Spec};
%% \draw (gcc |- 0,-3) node [align=center] {Model checker};
%% \draw (out |- 0,-3) node [align=center] {Proof \\ (written in $\mathcal{L}$)};

%% \end{tikzpicture}
%% }
%% \end{center}

%% \caption{Schematic diagram of {\sc synth}}
%% \label{fig:synth-dfd}
%% \end{figure}

%% Now that we have a procedure for solving a formula
%% in the synthesis fragment, we can build the full procedure for
%% deciding termination of a loop $L$.  This is given as Algorithm~\ref{alg:termination}.

%% \begin{algorithm}
%% {\small
%%  \caption{Bit-vector termination
%%  \label{alg:termination}}
%%  \vspace{-.5em}
%%  \begin{algorithmic}[1]
%%  \Statex
%% \Function{terminates}{$L$}
%%   %\Let{termination-spec}{$\exists W, R. \forall x, x' . \phi(W, R, x, x')$} \Comment{Formula {\bf [CT]}}
%%   %\Let{nonterm-spec}{$\exists N, C, x_0, \forall x. \psi(N, C, x_0, x)$} \Comment{Formula {\bf [SNT]}}
%%   \Statex  \Comment{Formula {\bf[GT]}:}
%%   \Let{spec}{$\exists P_T, P_N. \phi(P_T, x, x') \vee \psi(P_N, y)$}
%%   \Let{$(P_T, P_N)$}{{\sc Synthesise}(spec)}
%%   \If{decide($\exists x, x' . \lnot \phi(P_T, x, x'))$ = unsat}
%%     \State \Return{$L$ terminates}
%%   \ElsIf{decide($\exists x . \lnot \psi(P_N, x))$ = unsat}
%%     \State \Return{$L$ does not terminate}
%%   \EndIf
%% \EndFunction
%%  \end{algorithmic}
%%  }
%% \end{algorithm}

\section{Soundness, Completeness and Complexity}
The second-order SAT solver we are using is a semi-decision procedure -- for any
satisfiable formula, the solver is guaranteed to find a satisfying assignment to
all the second-order variables~\cite{kalashnikov}. 
%% Termination for bit-vector programs is known to be
%% PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the
%% expected results for soundness and completeness of our algorithm for
%% deciding it.  Our strategy for proving soundness and completeness
%% is first to prove that Algorithm~\ref{alg:cegis} (as instantiated
%% with \newC as a background logic) is a semi-decision
%% procedure.
In this section, we show that $\mathcal{L}$ is expressive enough
to compute (non-)termination proofs for every bit-vector program.
By using this result, %We will then combine this result with the fact that the second-order SAT solver is a semi-decision procedure to 
we then show that our technique terminates with a valid proof for every input program.

>>>>

\subsection{Solving the Second-Order SAT Formula}
In order to solve the second-order generalised formula {\bf [GT]}, we use the solver described in \cite{kalashnikov}.
%
%% \begin{theorem}
%%  \label{thm:synth-semicomplete}
%%  If the existential first-order theory used to express the specification $\sigma$ is decidable
%%  and the domain of inputs $X$ is finite,
%%  the second-order SAT decision procedure is semi-complete -- if a solution $(P, x_0) \models \sigma$
%%  exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
%%  satisfies the specification, the algorithm may not terminate.
%% \end{theorem}
%
%
This second-order SAT solver is a semi-decision procedure -- for any
satisfiable formula, the solver is guaranteed to find a satisfying assignment to
all the second-order variables.

In the context of our termination analysis, such a satisfying assignment returned by the solver 
represents either a proof of termination or non-termination, and
takes the form of an imperative program
written in the language $\mathcal{L}$.  An $\mathcal{L}$-program is a list
of instructions, each of which matches one of the patterns shown in
Figure~\ref{fig:l-language}.  An instruction has an opcode (such as
\verb|add| for addition) and one or more operands.  An operand is either a
constant, one of the program's inputs or the result of a previous
instruction.  The $\mathcal{L}$ language has various arithmetic and logical
operations, as well as basic branching in the form of the \verb|ite|
(if-then-else) instruction.

@@

\section{Soundness, Completeness and Complexity}
%% Termination for bit-vector programs is known to be
%% PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the
%% expected results for soundness and completeness of our algorithm for
%% deciding it.  Our strategy for proving soundness and completeness
%% is first to prove that Algorithm~\ref{alg:cegis} (as instantiated
%% with \newC as a background logic) is a semi-decision
%% procedure.
In this section, we show that $\mathcal{L}$ is expressive enough
to compute (non-)termination proofs for every bit-vector program.
By using this result, %We will then combine this result with the fact that the second-order SAT solver is a semi-decision procedure to 
we then show that our technique terminates with a valid proof for every input program.
